Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-03-11 ms.tex
fixed quataert discussion

Even if SN Ia are rare near the stagnation radius, they may cap the
SMBH accretion rate by periodically blowing gas out of the nucleus of
low mass galaxies.  Between successive supernovae, stars release a
gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$
interior to the Ia radius, which is gravitationally bound to the SMBH
by an energy $E_{\rm bind} \sim M_{\rm g}\sigma^{2}$.  Thus, according
to the above definitions, $E_{\rm Ia}/E_{\rm bind} \sim (v_{w}^{\rm
  Ia})^{2}/2\sigma^{2} \gtrsim 1$.  Hence, SN Ia are in principle
capable of dynamically clearing out gas from radii $\sim r_{\rm Ia}
\gtrsim r_{\rm s}$, at least for low mass BHs with $\sigma \ll v^{\rm
  Ia}_{w}$.  Thus, even in cases when heating is sufficiently weak
that the stagnation radius formally exceeds $r_{\rm Ia}$, the
accretion rate is limited to a maximum value,
\begin{eqnarray}
\frac{\dot{M}_{\rm Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{R_{\rm Ia}}{R_{\rm soi}}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp},
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
obtained substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}]) for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$ (eq.~[$\ref{eq:eddr_analytic}$]).  Note that $\dot{M}_{\rm Ia}$ usually exceeds the maximum accretion rate for a thermally stable flow $\dot{M}_{\rm TI}$ (eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow-out' from SNe Ia cannot alone prevent cooling instabilities in the nuclei of low mass galaxies.  Also note that Ia supernovae only cap the accretion rate to this maximum value if the gas feeding the black hole is supplied by local stellar mass loss: the gaseous mass can become much greater if the nucleus is fed externally, e.g. due to a galaxy merger or to cooling flows developing within the galaxy over much longer timescales (e.g.~\citealt{Ciotti&Ostriker07}).

@@

Figure \ref{fig:bh_growth} (top panel) shows with solid lines the
accretion rate as a function of black hole mass for the average star
formation heating, calculated from equation (\ref{eq:eddr_analytic})
using our results for the total wind heating
(Fig.~\ref{fig:mdot_mass}).  This average accretion rate increases
from $\dot{M} \sim 10^{-6}-10^{-5}M_{\odot}$ yr$^{-1}$ for low-mass
black holes ($M_{\bullet} \sim 10^{6}M_{\odot}$) to $\dot{M} \sim
10^{-4}M_{\odot}$ yr$^{-1}$ for $M_{\bullet} \sim 10^{9}M_{\sun}$.
These fall below the maximum thermally stable accretion rate,
$\dot{M}_{\rm TI}$ (plotted as yellow lines), except at the highest
black hole masses $\sim 10^{9}M_{\odot}$.  For these highest mass
black holes, the maximum accretion rate as set by gas blow-out from SN
Ia (green lines; eq.~[\ref{eq:eddr_Ia}]) could instead cap the
accretion rate below the thermally-unstable value, provided that the
SN Ia heating efficiency is sufficiently high that $v_{\rm w}^{\rm Ia}
\gg \sqrt{2}\sigma \sim 400$ km s$^{-1}$ (see discussion just
preceeding eq.~[\ref{eq:eddr_Ia}]).  This is only possible if the Ia
heating efficiency is higher $\epsilon_{\rm Ia} > 0.1$ than normally
assumed (\citealt{Sharma+14}).  For lower mass black holes, the
maximum accretion rate set by SN Ia is significantly higher and cannot
also prevent the development of thermally unstable flows,
i.e. $\dot{M}_{\rm Ia} \gg \dot{M}_{\rm TI}$.

Shown also for comparison in Figure \ref{fig:bh_growth} (top panel) is
the accretion rate onto SgrA* estimated by \citet{Quataert:2004a} and
the upper limit on $\dot{M}$ for the black hole in NGC3115 set by
\citealt{ShcherbakovWong+:2014a}.  These points are consistent with
our thermally-stable accretion rates within the expected dispersion
(see next section), although NGC3115 is in fact a low luminosity AGN.

\begin{figure}
\includegraphics[width=\columnwidth]{mdot_sfr.pdf}
\caption{\label{fig:bh_growth} {\it Top panel:} Critical values of the
  black hole accretion rate $\dot{M}$ in Eddington units as a function
  of black hole mass $M_{\bullet}$, shown separately for core
  ($\Gamma=0.1$, solid lines) and cusp galaxies ($\Gamma=0.8$,
  dot-dashed lines).  Black lines show the accretion rate calculated
  from equation (\ref{eq:eddr_analytic}) using our estimate of the gas
  heating rate provided at $z = 0$ by the average star formation
  histories for each galaxy mass (Fig.~\ref{fig:vwSources}).  Green
  lines show the maximum accretion set by Ia supernovae blow-out
  (\ref{eq:eddr_Ia}).  Red lines show the accretion rate so obtained
  if Compton heating alone was acting (eq.~[\ref{eq:MdotC}]).  Yellow
  lines show the maximum accretion rate for a thermally stable flow
  near the stagnation radius (eq.~[\ref{eq:Mdotmax}]).  We also show
  Eddington ratios inferred for the galactic center from
  \citet{Quataert:2004a} (black circle) and for NGC3115 from
  \citet{ShcherbakovWong+:2014a} (black diamonds). The top and bottom
  diamonds correspond to solution without and with conductivity
  respectively. Note that \citet{ShcherbakovWong+:2014a} do not
  include saturation of the thermal conductivity in their model {\it Bottom
    panel:} Growth times $\Mbh/\Mdot$ in units of the Hubble time
  $\th$ for each of the accretion rates shown in the top panel.}
\end{figure}

>>>>

Even if SN Ia are rare near the stagnation radius, they may cap the
SMBH accretion rate by periodically blowing gas out of the nucleus of
low mass galaxies.  Between successive supernovae, stars release a
gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$
interior to the Ia radius, which is gravitationally bound to the SMBH
by an energy $E_{\rm bind} \sim M_{\rm g}\sigma^{2}f^{2}$  {\bf BDM: update discussion}.  Thus, according
to the above definitions, $E_{\rm Ia}/E_{\rm bind} \sim (v_{w}^{\rm
  Ia})^{2}/2\sigma^{2} \gtrsim 1$.  Hence, SN Ia are in principle
capable of dynamically clearing out gas from radii $\sim r_{\rm Ia}
\gtrsim r_{\rm s}$, at least for low mass BHs with $\sigma \ll v^{\rm
  Ia}_{w}$.  Thus, even in cases when heating is sufficiently weak
that the stagnation radius formally exceeds $r_{\rm Ia}$, the
accretion rate is limited to a maximum value,
\begin{eqnarray}
\frac{\dot{M}_{\rm Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{R_{\rm Ia}}{R_{\rm soi}}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp},
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
obtained substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}]) for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$ (eq.~[$\ref{eq:eddr_analytic}$]).  Note that $\dot{M}_{\rm Ia}$ usually exceeds the maximum accretion rate for a thermally stable flow $\dot{M}_{\rm TI}$ (eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow-out' from SNe Ia cannot alone prevent cooling instabilities in the nuclei of low mass galaxies.  Also note that Ia supernovae only cap the accretion rate to this maximum value if the gas feeding the black hole is supplied by local stellar mass loss: the gaseous mass can become much greater if the nucleus is fed externally, e.g. due to a galaxy merger or to cooling flows developing within the galaxy over much longer timescales (e.g.~\citealt{Ciotti&Ostriker07}).

@@

Figure \ref{fig:bh_growth} (top panel) shows with solid lines the
accretion rate as a function of black hole mass for the average star
formation heating, calculated from equation (\ref{eq:eddr_analytic})
using our results for the total wind heating
(Fig.~\ref{fig:mdot_mass}).  This average accretion rate increases
from $\dot{M} \sim 10^{-6}-10^{-5}M_{\odot}$ yr$^{-1}$ for low-mass
black holes ($M_{\bullet} \sim 10^{6}M_{\odot}$) to $\dot{M} \sim
10^{-4}M_{\odot}$ yr$^{-1}$ for $M_{\bullet} \sim 10^{9}M_{\sun}$.
These fall below the maximum thermally stable accretion rate,
$\dot{M}_{\rm TI}$ (plotted as yellow lines), except at the highest
black hole masses $\sim 10^{9}M_{\odot}$.  For these highest mass
black holes, the maximum accretion rate as set by gas blow-out from SN
Ia (green lines; eq.~[\ref{eq:eddr_Ia}]) could instead cap the
accretion rate below the thermally-unstable value, provided that the
SN Ia heating efficiency is sufficiently high that $v_{\rm w}^{\rm Ia}
\gg \sqrt{2}\sigma \sim 400$ km s$^{-1}$ (see discussion just
preceeding eq.~[\ref{eq:eddr_Ia}]).  This is only possible if the Ia
heating efficiency is higher $\epsilon_{\rm Ia} > 0.1$ than normally
assumed (\citealt{Sharma+14}) and if the galaxy is a core.  For lower mass black holes, the
maximum accretion rate set by SN Ia is significantly higher and cannot
also prevent the development of thermally unstable flows,
i.e. $\dot{M}_{\rm Ia} \gg \dot{M}_{\rm TI}$.

Shown also for comparison in Figure \ref{fig:bh_growth} (top panel) is the accretion rate onto SgrA* calculated by \citet{Quataert:2004a}, as well as the range of $\dot{M}$ for the low-luminosity AGN NGC3115 from \citealt{ShcherbakovWong+:2014a} (who find a range of accretion rates depending on the assumed model for thermal conductivity, although they do not include saturation of the conductive flux).  The SgrA* rate exceeds our estimate of $\dot{M}$ due to the average star formation history for galaxies with the same mass black hole by almost two orders of magnitude.  This discrepency can be understood by the fact that star formation in the central parsec of the Galaxy is not in steady-state: feedback is instead dominated by the stellar winds from the ring of young massive stars of mass $\sim 10^{4}M_{\odot}$ and estimated age $\sim 10$ Myr (e.g., \citealt{Schodel+07}).  Such a star formation history is better described as a `burst' of stars, for which the value of $\eta \sim 100$ at $\tau_{\star} = 10^{7}$ yr (Fig.~\ref{fig:vwImp}) is a factor of $\sim 100$ times higher than the value of $\eta \sim 1$ predicted in the average SFH case (Fig.~\ref{fig:vwSources}, bottom panel).

\begin{figure}
\includegraphics[width=\columnwidth]{mdot_sfr.pdf}
\caption{\label{fig:bh_growth} {\it Top panel:} Critical values of the
  black hole accretion rate $\dot{M}$ in Eddington units as a function
  of black hole mass $M_{\bullet}$, shown separately for core
  ($\Gamma=0.1$, solid lines) and cusp galaxies ($\Gamma=0.8$,
  dot-dashed lines).  Black lines show the accretion rate calculated
  from equation (\ref{eq:eddr_analytic}) using our estimate of the gas
  heating rate provided at $z = 0$ by the average star formation
  histories for each galaxy mass (Fig.~\ref{fig:vwSources}).  Green
  lines show the maximum accretion set by Ia supernovae blow-out
  (\ref{eq:eddr_Ia}).  Red lines show the accretion rate so obtained
  if Compton heating alone was acting (eq.~[\ref{eq:MdotC}]).  Yellow
  lines show the maximum accretion rate for a thermally stable flow
  near the stagnation radius (eq.~[\ref{eq:Mdotmax}]).  We also show
  Eddington ratios inferred for the galactic center from
  \citet{Quataert:2004a} (black circle) and for NGC3115 from
  \citet{ShcherbakovWong+:2014a} (black diamonds).  {\bf BDM: turn quataert/sch into lines, add J1644+57 lines extending below 1e7 msun} {\it Bottom
    panel:} Growth times $\Mbh/\Mdot$ in units of the Hubble time
  $\th$ for each of the accretion rates shown in the top panel.}
\end{figure}
